\section{Related Work}
\label{sec:related_work}

\paragraph*{Aggregation in the provenance semiring framework}
The approach described here relies on that proposed
in~\cite{amsterdamer2011provenance}: \sql{HAVING} expressions result in
comparison operators between elements of a formal algebraic structure
combining semiring elements and scalar values (in
\cite{amsterdamer2011provenance}, a \emph{provenance semimodule}). But
\cite{amsterdamer2011provenance} stopped at producing these comparison
expressions, with no concrete computable semantics for these; probability
computation is also not considered.
The approach of \cite{amsterdamer2011provenance} is followed by \cite{pintor2025dbms}, in which a
DBMS-independent query rewriting system is proposed to produce such
expressions, stored as strings. The present work starts where these two
works end, by providing a semantics for these expressions.

\paragraph*{Negation in the provenance semiring framework}
Since provenance semirings \cite{green2007provenance,green2017provenance}
do not support non-monotone queries, \cite{geerts2010database} proposed
to extend them with a monus operator, yielding m-semirings. This approach
has limitations as some m-semirings do not respect axioms that may be
expected \cite{amsterdamer2011limitations}, in particular, as discussed
in Section~\ref{sec:preliminaries} distributivity of times over monus.
However, this modeling of negation has proved fruitful, being the basis
of a provenance for SPARQL queries
\cite{DBLP:journals/jacm/GeertsUKFC16}, of a semiring generalization of
Codd's theorem \cite{badia2025codd}, and of the implementation of the
\sql{EXCEPT} operator in ProvSQL \cite{sen2026provsql}. An alternative
modeling of negation is through
quotient semirings of polynomials with dual indeterminates
\cite{gradel2025provenance}, which is less practically implementable.

\paragraph*{Provenance systems with aggregation}
Two main provenance-aware database systems support aggregation.
ProvSQL \cite{sen2026provsql} is an extension of the PostgreSQL database
engine which rewrites queries over provenance-tracked relations
transparently and represents provenance annotations as nodes in a provenance
circuit. Its implementation of aggregation
closely follows
\cite{amsterdamer2011provenance} but, again, only supports terminal
aggregation -- this is the system we extend in this work. ProvSQL is
also a probabilistic database, which is relevant for this work.
GProM \cite{arab2018gprom} is a query-rewriting middleware supporting
different relational backends -- it does not support full semiring
semantics, but different forms of provenance can be computed and
aggregation is supported by joining the aggregation result with
provenance computed for each group; \sql{HAVING} is treated as a
downstream selection, meaning that GProM does not record how the
provenance value depends on whether the comparison holds. This in
particular precludes GProM from being used as a probabilistic database
engine.

\paragraph*{Aggregation in probabilistic databases}
The complexity of \sql{HAVING} queries over probabilistic databases has
been explored in \cite{re2009trichotomy}, a purely theoretical work that gives a full
characterization of
which aggregate conjunctive queries with a single \sql{HAVING} atom can
be evaluated or approximated in polynomial-time on a
probabilistic database; it is mostly orthogonal to the results in
Section~\ref{sec:algorithms} which explore which combinations of
semirings and aggregation operators can be evaluated in polynomial-time,
independently of the shape of the query.
With respect to probabilistic data, Proposition~\ref{prop:agg-cmp-poly-prob}
focuses on the tractability of the sole comparison operator, for
read-once provenance.

\cite{abiteboul2011capturing} consider aggregate queries over
probabilistic XML documents; they mainly study the complexity of
different operations over terminal aggregates, so there is no support of
\sql{HAVING} -- this is again not
implemented.

\cite{fink2012aggregation} is a very relevant work that also follows the
semiring construction of \cite{amsterdamer2011provenance} but extend it
for the specific purpose of probability evaluation, supporting
\sql{HAVING} operators in the process. This means in particular that
their semantics of \sql{HAVING} is specific to the probabilistic or
Boolean function case, in contrast with our proposed semantics valid in
arbitrary commutative m-semirings. Only aggregates over commutative
monoids are supported, as in
\cite{amsterdamer2011provenance}. The approach has been implemented in a
prototype version of SPROUT, unfortunately not publicly released.

In addition to ProvSQL~\cite{sen2026provsql} which was already discussed,
other probabilistic database systems have been released as research
prototypes:
MystiQ \cite{DBLP:conf/sigmod/BoulosDMMRS05},
Trio \cite{dblp:conf/vldb/agrawalbshnsw06},
Orion \cite{singh2008orion},
MayBMS \cite{huang2009maybms} and its SPROUT
\cite{DBLP:conf/icde/OlteanuHK09}\footnote{By SPROUT, here, we mean the publicly
  available version that is integrated within MayBMS or available from
  \url{https://www.cs.ox.ac.uk/projects/SPROUT/}; the SPROUT version
  from \cite{fink2012aggregation} was not publicly released, and
is discussed above.} query evaluator; none of them
is maintained anymore and they have either no support (MystiQ, MayBMS) or
only partial support (Trio~\cite{murthy2011aggregation}, Orion) for aggregation over
uncertain data; none of them supports an uncertain semantics for
\sql{HAVING}.

ProbLog \cite{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:conf/pkdd/DriesKMRBVR15} is a probabilistic
programming system, reminiscent of Prolog, that can be used to query
uncertain data by translating tuples to facts and queries to rules. It
does support aggregation, and filtering on aggregation results is also
supported -- marginal probabilities can be obtained through ground
computation and weighted model counting; evaluation in other
(m-)semirings is not possible. Importantly, \cite{sen2026provsql} has
shown that ProbLog which relies on in-memory processing of rules, cannot
handle databases of real-world sizes, in contrast with dedicated
probabilistic databases.

\paragraph*{Aggregation over compact representations}

Finally, \cite{bakibayev2013aggregation} extends a factorized database
engine with aggregation and \sql{HAVING} clauses. Though no provenance is
generated and the setting is deterministic, there is some overlap in
techniques with those necessary to track provenance. In particular, one
possible direction for future work would be to investigate whether such a
factorized representation could be used to compress the possible worlds
enumerated over in our semantics.
